---
title: 'Data Visualization in R'
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

The `Loblolly` data in `R` has several variables pertaining to growth records 
for Loblolly pines, a type of pine tree native to the Southeastern United States.
Load this data in `R` and examine the help file with the following commands:

```{r eval=FALSE}
data("Loblolly")
?Loblolly
```

1.  What are the variables in this dataset? Are they numeric or categorical?

## Boxplots

In order to get more comfortable examining model assumptions, we'd like to get
familiar with `R`'s plotting capailibites. We will start by examining how `height`
varies across different levels of `seed`. Since the `seed` variable has 14 levels,
we will ask `R` for a subset of the data that includes only seeds 329, 315, and 305. 

The following command creates this subset by taking `Loblolly` *such that* `Seed` 
is 329 *or* `Seed` is 315 *or* `Seed` is 305. The `droplevels` command cleans up 
the subsetted data so that it will plot nicely. 

```{r eval=FALSE}
subset <- Loblolly[Loblolly$Seed == 329 | Loblolly$Seed == 315 | Loblolly$Seed == 305,]
subset <- droplevels(subset)
```

To examine graphically how `height` varies across different levels of `seed` 
(in our subset of the data), we will start with a boxplot. Remember that we can
use a `~` as "by". That is, we want a boxplot of `height` *by* levels of `seed`.
Recall also that we use the dollar sign `$` to tell `R` that we want a particular
variable from a dataset, i.e., dataset$variable.

```{r eval=FALSE}
boxplot(subset$height ~ subset$Seed)
```
    
This plot is a nice start, but it may look somewhat incomplete. It's missing a 
title and could stand to have cleaner axis labels. The following command adds
a `main` title and axis labels `xlab` and `ylab`:

```{r eval=FALSE}
boxplot(subset$height ~ subset$Seed, 
        main = "Boxplot of Tree Height by Seed on Subsetted Data",
        xlab = "Seed", ylab="Height (ft)")
```

2.  Do you think that `height` differs between different values of `seed`? 

3.  Do the three `height` groups look normally distributed? 

## Histograms

While boxplots are a convenient way to make side-by-side comparisons, it can
be difficult to conclusively answer questions like the one posed in Exercise 3. 
Histograms provide a much more straightforward way to examine the shape of a 
distribution. The following command creates a basic histogram for the `Loblolly`
`height` variable:

```{r eval=FALSE}
hist(Loblolly$height)
```

We would again like to include a better title and new axis labels. Fortunately,
`R`'s functions for plotting all use the same approach!

```{r eval=FALSE}
hist(Loblolly$height,
     main = "Histogram of Loblolly Pine Heights",
     xlab = "Height (ft)", ylab="Frequency")
```

4.  Do the Loblolly pine heights appear to be normally distributed?

Previously, we wanted information on normality for a subset of the data,
using only seeds 329, 315, and 305. We can build individual histograms for
these data. Recall that the square brackets can be read as "such that".

```{r eval=FALSE}
hist(Loblolly$height[Loblolly$Seed == 329],
     main = "Histogram of Pine Heights for Seed 329",
     xlab = "Height (ft)", ylab="Frequency")
```

5.  Create histograms of the pine heights for the other two seeds, 315 and 305.
    For each seed, decide whether it is reasonable to assume that the heights 
    are normally distributed. 
    
## Scatterplots and Regression Lines

We may run into some problems with this data because there are only 6 
observations per seed! It may be more reasonable to compare `age` and `height` 
of trees in the complete data.

6.  How many observations are there for `height`? For `age`?

If we want to examine the relationship between `age` and `height`, it is 
reasonable to think that we would be interested in using `height` to 
predict `age`. (If we walking through a forest of Loblolly pine trees, it will 
be much easier to get a tree's height than its age!) 

7. In this setting, which is the predictor variable? Which is the response?

Using this predictor/response variable setting, we want to look at a scatterplot
of the data to get an idea of whether there might be any correlation between the 
two. The following command creates this scatterplot, complete with a title and 
reasonable axis labels. 

```{r eval=FALSE}
plot(x = Loblolly$height, y = Loblolly$age,
     main = "Scatterplot of Age vs Height",
     xlab = "Height (feet)", ylab = "Age (Years)")
```

8.  Is there evidence of a linear relationship between tree age and tree height?
    Without doing any math or using the computer, take a guess as to what the 
    correlation might be for these two variables.
    
We'll spend some time on regression next week, but for now the regression line is
\[
  \hat{y} = 0.7574 + 0.3783 x
\]
We can include this in our plot using the function `abline`. This function adds a 
line to an existing plot in `R`. The name "abline" refers to the way that lines are
written in many an algebra class: $y = a + bx$. Include the regression line in
your scatterplot by adding the following line of code right under the previous plot 
function:

```{r eval=FALSE}
abline(a = 0.7574, b = 0.3783)
```

9.  Predict the age of a tree that is 20 feet tall.

* * *
## On your own

-   In class, we saw plots of the `faithful` dataset. We thought about using the waiting
    time to predict eruption duration. Now, we will think about using eruption duration
    to predict how long we need to wait before seeing another eruption. Load the data
    using the command `data("faithful")`.

    **a.** What is the response variable? What is the predictor variable?
    
    **b.** Create a scatterplot of this data with the predictor and response variables
    on the appropriate axes. Make sure to include appropriate axis labels and title.
    
    **c.** The regression line for this data is 
    \[
      \hat{y} = 33.47 + 10.73x
    \]
    Add this line to your scatterplot.

-   The `ToothGrowth` dataset in `R` gives information about the effect of Vitamin C
    on tooth grown in Guinea Pigs. (Use the command `?ToothGrowth` for more information
    on the data and each individual variable.) Load this data into `R` using the 
    command `data("ToothGrowth")`.

    **a.** Create boxplots of the guinea pigs' tooth length by the factor levels for 
    supplement type *and* dosage. Be sure to include an appropriate title and 
    axis labels to match your boxplots. *Hint*: you can create factor levels in a 
    formula by using factorA*factorB.

    **b.** Based on your boxplots, do you think there are differences among the six
    treatments?
    
    **c.** Create a histogram of the guinea pigs' tooth lengths. Is it reasonable to 
    assume normality?

    **d.** Create an interaction plot for this data. Use the command `?interaction.plot` 
    to examine the function call to create an interaction plot in `R`. You can 
    ignore most of the inputs for this function, but make sure to pay attention 
    to `x.factor`, `trace.factor`, and `response`. Make sure to include appropriate 
    axis labels, title, and trace label (`trace.label = "SOME TEXT"`).
    *Hint*:  The "trace" refers to the variable that is split into multiple lines. 
    The main title and axis labels can be set in the same way as in the other plots.

<div id="license">
This lab was written by Lauren Cappiello for STAT 100B at the University of 
California, Riverside using the RMarkdown Lab style file from OpenIntro.
</div>