---
title: 'Inference for numerical data: t-tests'
output:
  html_document:
    css: /Users/cappiello/Documents/stat100b/oiLabs-base-R-2.0.0/lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

## North Carolina births

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

## Exploratory analysis

Load the `nc` data set into our workspace.

```{r load-data, eval=FALSE}
download.file("http://www.openintro.org/stat/data/nc.RData", destfile = "nc.RData")
load("nc.RData")
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | -----------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

1.  What are the cases in this data set? How many cases are there in our sample?

As a first step in the analysis, we should consider summaries of the data. This
can be done using the `summary` command:

```{r summary, eval=FALSE}
summary(nc)
```

As you review the variable summaries, consider which variables are categorical 
and which are numerical. For numerical variables, are there outliers? If you 
aren't sure or want to take a closer look at the data, make a graph.

Consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

2.  Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following function to 
split the `weight` variable into the `habit` groups, then take the mean of each
using the `mean` function.

```{r by-means, eval=FALSE}
by(nc$weight, nc$habit, mean)
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test
.

## T Tests

3.  Check if the conditions necessary for inference are satisfied. Note that 
you will need to obtain sample sizes to check the conditions. You can compute 
the group size using the same `by` command above but replacing `mean` with 
`length`.

4.  Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.

Next, we introduce a new function, `t.test`, that we will use for conducting
hypothesis tests and constructing confidence intervals. 

```{r eval=FALSE, tidy=FALSE}
t.test(nc$weight ~ nc$habit, alternative = "two.sided", mu = 0, 
       paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```

The output for this test includes the test statistic `t`, the degrees of
freedom, `df` and the `p-value`. It also shows the alternative hypothesis,
a confidence interval, and the means for each group. Did you notice that
the degrees of freedom look strange? The function `t.test` uses what is 
called the "Welch modification" to estimate the degrees of freedom for 
unequal variances. The details of this are outside of the scope of this
course. 

Let's pause for a moment to really break down the anatomy of this function.

```{r eval=FALSE}
?t.test
```


There are two ways to approach `t.test`. The first is with `x` and (optionally) 
`y`. Using only `x`, we can test whether the mean of `x` is equal to some
null value mu. Using `x` and `y`, we can test whether the difference in means
for `x` and `y` is equal to mu. We can also use `formula`, which is what
we've done above. Here, we write `y~x`, which in this case means "y by x".
That is, `y` is the response variable: `nc$weight` and `x` is the explanatory 
variable that splits the data into two groups, `smokers` and `nonsmokers`:
`nc$habit`. The next argument, `alternative` specifies the alternative 
hypothesis, `mu` specificies the null value, `paired` specifies a paired 
samples setting, and `var.equal` specifies whether the variances are equal
for the two groups. Finally, `conf.level` specifies the confidence level. 

5.  Change the `var.equal` argument to `TRUE` to retest this hypothesis under
the assumption that the variances are equal between the groups for weights of 
babies born to smoking and non-smoking mothers. Carefully report your
conclusions. 

* * *

## On your own

-   Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) and interpret it in context. 

-   Calculate a new confidence interval for the same parameter at the 90% 
confidence level. 

-   Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

-   Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.

-   Pick a pair of numerical and categorical variables and come up with a 
research question evaluating the relationship between these variables. 
Formulate the question in a way that it can be answered using a hypothesis test
and/or a confidence interval. Answer your question using the `t.test` 
function, report the statistical results, and also provide an explanation in 
plain language.

<div id="license">
Note: This lab is derivative of an OpenIntro lab, released under a Creative Commons Attribution-ShareAlike 3.0 Unported. The original OpenIntro lab may be found [here](http://htmlpreview.github.io/?https://github.com/andrewpbray/oiLabs-base-R/blob/master/inf_for_numerical_data/inf_for_numerical_data.html). 
</div>