%Section 11.1 from Mendenhall, Beaver, & Beaver
\begin{frame}{Experimental Design}
    You know a lot about experimental design already, but we want to contextualize some of those concepts.
\end{frame}

\begin{frame}{Observational Studies}
    For an \textbf{observational study},
    \begin{itemize}
        \item A \textbf{sampling plan} refers to the way a sample is selected.
        \item Data is observed rather than produced.
        \item Example: sample surveys.
    \end{itemize}
\end{frame}

\begin{frame}{Experiments}
    For an \textbf{experiment}, 
    \begin{itemize}
        \item \textbf{Experimental design} refers to the way a sample is selected.
        \item One or more conditions are imposed by the researcher.
    \end{itemize}
\end{frame}

\begin{frame}{Experimental Design Terms}
    \begin{itemize}
        \item An \textbf{experimental unit} is the object on which a measurement is taken.
        \item A \textbf{factor} is an independent variable whose values are controlled/varied by the experimenter. 
        \item A \textbf{level} is the intensity (value) of a factor.
        \item A \textbf{treatment} is a specific combination of factor levels.
        \item The \textbf{response} is the variable being measured. 
    \end{itemize}
\end{frame}

\begin{frame}{Example}
    \begin{itemize}
        \item A group of people is randomly divided into an experimental and a control group. 
        \item The control group is given an aptitude test after having eaten a full breakfast. 
        \item The experimental group is given the same test without having eaten any breakfast. 
    \end{itemize}
    What are the factors, levels, and treatments in this experiment?
\end{frame}

\begin{frame}{Example}
    \begin{itemize}
        \item Suppose that the experimenter in the previous example began by randomly selecting 20 men and 20 women for the experiment. 
        \item These two groups were then randomly divided into 10 each for the experimental and control groups. 
    \end{itemize}
    What are the factors, levels, and treatments in this experiment?
\end{frame}
